An Evolutionary Algorithm for the Multiobjective Risk-Equity Constrained Routing Problem

### 2011-10-17 code/variatorHazmatJava/DAM_Paper_EvolOptim_HazmatRouting/hazmatRouting_EvolOptim.tex
add the report in the DAM format

\documentclass{ctwart}
\usepackage{amssymb}
\usepackage{color}
% \usepackage{dsfont}
\usepackage[numbers,sort&compress]{natbib}

\newcommand{\TODO}[1]{{\color{red}TODO: #1}}
\newcommand{\TODON}[1]{{\color{blue}TODO: #1}}
\newcommand{\COMMENTD}[1]{{\color{green}Dimo: #1}}
\newcommand{\COMMENTN}[1]{{\color{green}Nora: #1}}

\newcommand{\mypara}[1]{{\bfseries #1: }}

@@

@@

\author{Nora Touati-Moungla and Dimo Brockhoff}

\address{Laboratoire d'Informatique, \'{E}cole Polytechnique, 91128 Palaiseau Cedex, France}

@@

\end{frontmatter}

\section{Introduction}

@@

Let the transportation network be represented as a directed graph $G = (N, A)$, with $N$ being the set of nodes and $A$ the set of arcs. Let $C$ be the set of commodities, given as a set of point-to-point demands to transport a certain amount of hazmats. For any commodity $c\in C$, let $s^c$ and $t^c$ be the source node and the destination node respectively, and let $V^c$ be the number of available trucks for the transportation of commodity $c$. Each commodity is associated with a unique type of hazmat. We assume that the risk is computed on each arc of the network and is proportional to the number of trucks traversing such an arc. We consider a set $Q$ of regions, and we define $r_{ij}^{cq}$ as the risk imposed on region $q \in Q$ when the arc $(i,j) \in A$ is used by a truck for the transportation of hazmat of type $c$. We remark that we employ a notion of {\em spread risk}, in that an accidental event on arc $(i,j)$ within region $q\in Q$ may strongly affect another region $q'\in Q$. With each arc $(i,j) \in A$ a cost $c_{ij}^c$ is associated, involved by the travel of a truck of commodity $c$ on this arc.

The problem of transporting hazmat is multiobjective in nature: one usually wants to minimize the \textit{total cost} of transportation, the \textit{total risk} of transportation imposed on all regions and the {\em distributed risk}, which can be defined as a measure of risk that is shared among different regions. More specifically, for a given solution, each region $q\in Q$ will be affected by a risk $\omega_q$ which depends on the transportation patterns in all other regions. The third objective will then be $\max_{q\in Q} \omega_q$, and has to be minimized.

We introduce the integer variable $y_{ij}^c$ for the number of trucks that use arc $(i,j)$ for transporting commodity $c$. We assume a fixed number of trucks and that all trucks have the same load. The proposed model is defined as follows:
\begin{eqnarray}
   \min  & \sum_{c \in C} \sum_{(i,j) \in A} c_{ij}^{c} y_{ij}^c & \label{eq:obj1}\\
   \min  & \sum_{c \in C} \sum_{q\in Q} \sum_{(i,j) \in A} r_{ij}^{cq} y_{ij}^c & \label{eq:obj2}\\
   \min  & \max_{q\in Q} \big\{ \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^{cq} y_{ij}^c \big\} & \label{eq:obj3}\\
    s.t. & \sum_{j \in \delta^+(i)} y_{ij}^c -  \sum_{j \in \delta^-(i)} y_{ji}^c = q_i^c\quad         & \forall i \in N,  c \in  C \label{eq:flow} \\
         & y_{ij}^c \in \{0, 1, 2, \ldots\}                                                            & \forall (i,j)\in A, c\in C\label{eq:integerY}.
\end{eqnarray}

The first objective in (\ref{eq:obj1}) is a cost function and is to be minimized. The second objective, given by (\ref{eq:obj2}), minimizes the total risk on all regions and objective (\ref{eq:obj3}) minimizes the maximum risk imposed on all regions. Constraints (\ref{eq:flow}) are conservation constraints, where $\delta^-(i) = \{j \in N: (j,i)\in A\}$ and $\delta^+(i) = \{j \in N: (i,j)\in A\}$ are the direct successors and predecessors of node $i$, and $q_i^c=V^c$ if  $i=s^c$, $q_i^c=-V^c$ if  $i=t^c$ and $q_i^c=0$ otherwise.

@@

\mypara{Representation}
We choose a variable length representation as it has been theoretically shown to be a good choice for multiobjective shortest paths problems \citep{horo2009a}: A solution is thereby represented by a list of paths of variable lengths with one path per truck. For the moment, we consider a fixed amount of trucks for each commodity and therefore a fixed number of paths through the network. In order to have every variable length path represent an uninterrupted path from source to destination at any time (see the constraints in (\ref{eq:flow})), we ensure all paths to always start with the source $s^c$ for the corresponding commodity $c$, ensure with the variation operator that all neighbored vertices in the path are connected by an arc, and complete each path by the shortest path between the path's actual end node and the commodity's destination node $t^c$.

\mypara{Initialization}
Initially, we start with a single solution where the paths $p$ for all trucks are empty ($p=(s^c)$). This corresponds to the situation where all trucks choose the shortest $s^c$--$t^c$ path for their assigned commodity---implying the smallest possible overall cost but a high risk along the used route(s). Nevertheless, the initial solution is already Pareto-optimal and is expected to be a good starting point for the algorithm.

\mypara{Variation}
As mutation operator, we suggest to shorten or lengthen the path of one or several trucks. In order to generate a new solution $s'$ from $s$, for each truck path, we draw a binary value $b\in\{0,1\}$ uniformly at random and create the new path $p'$ from the old one $p=(v_1=s^c,v_2,\ldots, v_l)$ as in \citep{horo2009a}:
\begin{itemize}
	\item if $b = 0$ and $l=:\textnormal{length}(p)\geq 2$, set $p' = (s^c, \ldots, v_{l-1})$	
	\item if $b = 1$ and $|V_{\textnormal{\scriptsize rem}}=\{v \in V \,|\, (v_{l}, v) \in A\}| \neq\emptyset$, choose $v_{l+1}$ from $V_{\textnormal{\scriptsize rem}}$ uniformly at random and set $p' = (v_0, \ldots, v_l, v_{l+1})$.
	\item otherwise, use the same path $p$ also in the new solution $s'$.
\end{itemize}

@@

\begin{thebibliography}{00}
\providecommand{\natexlab}[1]{#1}
\providecommand{\url}[1]{\texttt{#1}}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi: #1}\else
  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

@@

@@

>>>>

%% This is file `elsarticle-template-1a-num.tex',
%%
%% Copyright 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%%
%% $Id: elsarticle-template-1a-num.tex 151 2009-10-08 05:18:25Z rishi $
%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1a-num.tex $
%%
\documentclass[preprint,12pt]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
%% \usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
%% \usepackage{lineno}

%% natbib.sty is loaded by default. However, natbib options can be
%% provided with \biboptions{...} command. Following options are
%% valid:

%%   round  -  round parentheses are used (default)
%%   square -  square brackets are used   [option]
%%   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option>
%%   semicolon  -  multiple citations separated by semi-colon
%%   colon  - same as semicolon, an earlier confusion
%%   comma  -  separated by comma
%%   numbers-  selects numerical citations
%%   super  -  numerical citations as superscripts
%%   sort   -  sorts multiple citations according to order in ref. list
%%   sort&compress   -  like sort, but also compresses numerical citations
%%   compress - compresses without sorting
%%
%% \biboptions{comma,round}

% \biboptions{}

\journal{DAM}

@@

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

@@

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

\author{Dimo Brockhoff and Nora Touati-Moungla}

\address{Laboratoire d'Informatique, \'{E}cole Polytechnique, 91128 Palaiseau Cedex, France\\
\{Brockhoff, touati\}@lix.polytechnique.fr}

\begin{abstract}

\end{abstract}

@@

\end{frontmatter}
\section{Introduction}

@@

Let the transportation network be represented as a directed graph $G = (N, A)$, with $N$ being the set of nodes and $A$ the set of arcs. Let $C$ be the set of commodities, given as a set of point-to-point demands to transport a certain amount of hazmats. For any commodity $c\in C$, let $s^c$ and $t^c$ be the source node and the destination node respectively, and let $d^c$ be the amount of hazmat to be shipped from $s^c$ to $t^c$. We assume that the risk is computed on each arc of the network and is proportional to the arc length and the population density around this arc, and we define $r_{ij}^c$ as the risk imposed on the population when the arc $(i, j) \in A$ is used for the transportation of one unit of hazardous materials of type $c$. With each arc $(i,j) \in A$ a cost $c_{ij}$ is associated, equivalent to the length of arc $(i,j)$.

\subsection{Multiple objective functions}
\label{SS_MOF}

The problem of transporting hazmat is multiobjective in nature: one usually wants to minimize the \textit{total cost} of transportation, the \textit{total risk} of transportation and the {\em distributed risk}, which can be defined as a measure of risk that is shared among different arcs or different paths.

\subsection{An optimization model}
\label{SS_AOM}

We introduce a flow variable $f_{ij}^c$ defining the portion of commodity $c$ being transported on arc $(i,j)$. These variables are subject to flow conservation constraints:

$$\sum_{j \in \delta^+(i)} f_{ij}^c -  \sum_{j \in \delta^-(i)} f_{ji}^c = b_i^c, \forall i \in N,  c \in  C$$

where $\delta^-(i)$ and $\delta^+(i)$ are the forward and backward star of $i$, i.e., $\delta^-(i) = \{j \in N: (j,i)\in A\}$ and $\delta^+(i) = \{j \in N: (i,j)\in A\}$, and $b_i^c=\left\{
\begin{array}{rl}
                     1 & \textrm{ if } i=s^c\\
                    -1 & \textrm{ if } i=t^c\\
                     0 & \textrm{ otherwise.}\\
\end{array}\right.$

The first objective is a cost function given as follows:

$$\min \sum_{c \in C} \sum_{(i,j) \in A} c_{ij} f_{ij}^c$$

The second objective is the minimization of the total risk

$$
\min \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^c f_{ij}^c
$$

We define the risk $w_{ij}$ imposed on the population around the arc $(i,j)$ as a linear combination of the flow variables:

$$w_{ij} = \sum_{c \in C} r_{ij}^c f_{ij}^c$$

and add a new variable $z = \max_{(i,j) \in A} w_{ij}$, which therefore is subject to the constraints

$$
z \geq \sum_{c \in C} r_{ij}^c f_{ij}^c, \forall (i,j) \in A
$$

% We define the risk $w_p$ imposed on the population around the path $p$ as a linear combination of the flow variables: 
% 
% $$w_p = \sum_{c \in C} \sum_{(i,j) \in p} r_{ij}^c f_{ij}^c$$
% 
% and add a new variable $y = \max_{p \in \mathcal{P}} w_p$, which therefore is subject to the constraints
% 
% $$
% y \geq \sum_{c \in C} \sum_{(i,j) \in p} r_{ij}^c f_{ij}^c, \forall p \in \mathcal{P}
% $$
% 
% where $\mathcal{P}$ is the set of paths.

% 
% 
% 
% The third and fourth objectives will then be:
% $$\min \max_{(i,j)\in A} r_{ij} f_{ij}^c$$, and 
% $$\min \max_{p \in \mathcal{P}} \sum_{(i,j) \in p} r_{ij} f_{ij} $$ 
% 
% where

% We introduce the integer variable $y_{ij}^c$ for the number of trucks that use arc $(i,j)$ for transporting commodity $c$. We assume a fixed number of trucks and that all trucks have the same load.

The proposed model is defined as follows:
\begin{eqnarray}
   \min  &  \sum_{c \in C} \sum_{(i,j) \in A} c_{ij} f_{ij}^c & \label{eq:obj1}\\
   \min  &  \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^c f_{ij}^c & \label{eq:obj2}\\
   \min  & z & \label{eq:obj3}\\
   %\min  & y & \label{eq:obj4}\\
    s.t. & \sum_{j \in \delta^+(i)} f_{ij}^c -  \sum_{j \in \delta^-(i)} f_{ji}^c = d_i^c\quad         & \forall i \in N,  c \in  C \label{eq:flow} \\
    & z \geq \sum_{c \in C} r_{ij}^c f_{ij}^c & \forall (i,j) \in A \\
%     & y \geq \sum_{c \in C} \sum_{(i,j) \in p} r_{ij}^c f_{ij}^c & \forall p \in \mathcal{P}
%          & y_{ij}^c \in \{0, 1, 2, \ldots\}                                                            & \forall (i,j)\in A, c\in C\label{eq:integerY}.
\end{eqnarray}

% The first objective in (\ref{eq:obj1}) is a cost function and is to be minimized. The second objective, given by (\ref{eq:obj2}), minimizes the total risk on all regions and objective (\ref{eq:obj3}) minimizes the maximum risk imposed on all regions. Constraints (\ref{eq:flow}) are conservation constraints, where $\delta^-(i) = \{j \in N: (j,i)\in A\}$ and $\delta^+(i) = \{j \in N: (i,j)\in A\}$ are the direct successors and predecessors of node $i$, and $q_i^c=V^c$ if  $i=s^c$, $q_i^c=-V^c$ if  $i=t^c$ and $q_i^c=0$ otherwise.

% \begin{Rq}
% The number of constaints $nbC = |N| + |A| + |\mathcal{P}|$ is exponontial, so the classical optimization softwares such as CPLEX can not be used.
% \end{Rq}

@@

\subsection{Representation}
We choose a variable length representation as it has been theoretically shown to be a good choice for multiobjective shortest paths problems \citep{horo2009a}: A solution is thereby represented by a list of paths of variable lengths with one path per truck. For the moment, we consider a fixed amount of trucks for each commodity and therefore a fixed number of paths through the network. In order to have every variable length path represent an uninterrupted path from source to destination at any time (see the constraints in (\ref{eq:flow})), we ensure all paths to always start with the source $s^c$ for the corresponding commodity $c$, ensure with the variation operator that all neighbored vertices in the path are connected by an arc, and complete each path by the shortest path between the path's actual end node and the commodity's destination node $t^c$.

\subsection{Initialization}
Initially, we start with a single solution where the paths $p$ for all trucks are empty ($p=(s^c)$). This corresponds to the situation where all trucks choose the shortest $s^c$--$t^c$ path for their assigned commodity---implying the smallest possible overall cost but a high risk along the used route(s). Nevertheless, the initial solution is already Pareto-optimal and is expected to be a good starting point for the algorithm.

\subsection{Variation}
As mutation operator, we suggest to shorten or lengthen the path of one or several trucks. In order to generate a new solution $s'$ from $s$, for each truck path, we draw a binary value $b\in\{0,1\}$ uniformly at random and create the new path $p'$ from the old one $p=(v_1=s^c,v_2,\ldots, v_l)$ as in \citep{horo2009a}:
\begin{itemize}
	\item if $b = 0$ and $l=:\textnormal{length}(p)\geq 2$, set $p' = (s^c, \ldots, v_{l-1})$	
	\item if $b = 1$ and $|V_{\textnormal{\scriptsize rem}}=\{v \in V \,|\, (v_{l}, v) \in A\}| \neq\emptyset$, choose $v_{l+1}$ from $V_{\textnormal{\scriptsize rem}}$ uniformly at random and set $p' = (v_0, \ldots, v_l, v_{l+1})$.
	\item otherwise, use the same path $p$ also in the new solution $s'$.
\end{itemize}

\section{Experimental analysis}

We concentrated our analysis on a real-world case study \cite{bianco09}. We considered the road network of the Lazio region (located in the middle of Italy), and, in particular, its main transport roads. The network is composed of 311 nodes and 441 links. The risk $r_{ij}^c, \forall (i,j) \in A, c \in C$ is evaluated as the societal risk computed as the number of people living inside the exposure zone around link $(i,j)$ (whose size depends on the hazmat type $c$ times the accident probability involving the vehicle. We considered from 2 to 10 origin-destination pairs on the network, each one associated with a number of commodities ranging from 1 to 3, for an overall number of shipments (commodities) between 2 and 30 (we recall that in our model each commodity is associated with one origin-destination pair). With each one of the latter scenarios we associated 10 instances; each instance has been generated by assigning uniformly at random a demand from 100 to 1000 tons to each shipment.

@@

\begin{thebibliography}{00}
% \providecommand{\natexlab}[1]{#1}
% \providecommand{\url}[1]{\texttt{#1}}
% \expandafter\ifx\csname urlstyle\endcsname\relax
%   \providecommand{\doi}[1]{doi: #1}\else
%   \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

@@

\bibitem[bianco(2009)]{bianco09}
L. Bianco and M. Caramia and S. Giordani.
\newblock A bilevel flow model for hazmat transportation network design.
\newblock Transportation research. Part C, Emerging technologies, 17(2):175--196, 2009.

@@

%%
%% End of file `elsarticle-template-1a-num.tex'.
